Unity in diversity: Children from 17 diverse communities process gaze in similar ways

### 2024-01-30 paper/paper.Rmd
update paper

```{r fig1, include = T, fig.align = "center", fig.cap = "(A) Data collection sites. Points show approximate geographical location of the data collection sites, coloring shows the sample sizes. (B) Screenshot from the task. The scene depicts the choice phase in a test trial. Participants had to use the gaze of the agent to locate the balloon and click on the hedge where they thought the balloon was. Agents, audio recordings and backgrounds were adpated to each cultural setting. (C) Graphical illustration of the cognitive model. Individuals infer the target of an agent's attention by estimating a vector based on the position of the pupils within the eyes. This process is noisy, illustrated by the different vectors. Individuals differ in their level of precision (indicated by alpha). Regardless of the level of precision, the further out the target lands, the less precise the model predicts individuals to be.", out.width="75%"}
knitr::include_graphics("../visuals/fig1.png")
```

@@

As the first step, we investigated developmental gains. Across all 17 communities, we found a substantial increase in average levels of precision with age (fixed effect: $\beta$ = `r post_fix%>%filter(term == "ageinyears")%>%pull(Estimate)`, 95% HDI (`r post_fix%>%filter(term == "ageinyears")%>%pull(Q2.5)` - `r post_fix%>%filter(term == "ageinyears")%>%pull(Q97.5)`); range of community-level effects: $\beta_{min}$ = `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(lci)` - `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(uci)`) to $\beta_{max}$ = `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(lci)` - `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(uci)`). There were marked differences between communities: highly market-integrated communities around the globe showed higher levels of precision compared to less market-integrated communities. However, we believe that these results are better explained by exposure to technology than market integration. We discuss them in more detail below. Differences between communities were small compared to differences between individuals: communities did not form homogeneous clusters but largely overlapping distributions in that some individuals from communities with a lower average level of precision performed better compared to some individuals from a setting with a very high average level of precision. Furthermore, in all communities, some 4-year-olds outperformed children two years older than them (see Fig. \@ref(fig:fig2)A).

@@

The processing signature was clearly visible across all 17 communities. Precision decreased when the agent looked at locations further away from the center (fixed effect: $\beta$ = `r post_fix%>%filter(term == "targetcentralityx")%>%pull(Estimate)`, 95% HDI (`r post_fix%>%filter(term == "targetcentralitx")%>%pull(Q2.5)` - `r post_fix%>%filter(term == "targetcentralityx")%>%pull(Q97.5)`); range of community-level effects: $\beta_{min}$ = `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(lci)` - `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(uci)`) to $\beta_{max}$ = `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(lci)` - `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(uci)`). Visualization of the data showed the predicted u-shaped pattern in all communities (see Fig. \@ref(fig:fig2)B). These results indicate a universal cognitive process used by children in all communities. There is, however, an alternative way in which the u = shaped pattern might arise: it would also arise when participants ignored the agent's gaze and instead always selected the middle of the screen (center bias). To rule out this alternative explanation, we directly compared three cognitive models that made different assumptions about how participants' responses were generated: the focal vector-based gaze estimation model described above, a center-bias model where participants always select the center, and a random guessing model where participants select random locations. For every community, we found overwhelming support for the gaze estimation model (min $BF_{10}$ < 1000 for comparisons with both alternative models). Taken together, children from all 17 communities processed gaze in similar ways.

@@

On an individual level, other factors likely generated individual differences because there was substantial variation even in communities where almost all children had access to touchscreens. Previous work suggested that social cognition develops in social interaction [@barresi1996intentional; @perner1994theory; @rakoczy2022foundations], in particular, with older siblings [@lewis1996social; @ruffman1998older]. We found tentative support for this idea in the form of a small but consistent effect in the direction that participants living together with more children older than themselves had slightly higher levels of precision ($\beta$ = `r pred_est%>%filter(term == "older_children")%>%pull(Estimate)`, SE = `r pred_est%>%filter(term == "older_children")%>%pull(Est.Error)`, 95% CrI = `r pred_est%>%filter(term == "older_children")%>%pull(Q2.5)` - `r pred_est%>%filter(term == "older_children")%>%pull(Q97.5)`).

Following and understanding gaze is a foundational building block of human social cognition [@tomasello2007reliance; @scaife1975capacity]. A substantial body of work has explored the developmental onset of gaze following in a few selected cultural settings [@tang2023slow; @gredeback2010development; @byers2021development; @hernik2019infant; @callaghan2011early]. The data reported here provides evidence that children all over the world process gaze in the same way. Key performance signatures of a model that sees gaze following as a form of social vector estimation were found in the data of all 17 distinct cultural communities. The cognitive processes underlying gaze following are thus likely to be rooted in humans' evolved cognitive architecture, which is -- presumably -- later refined during social interaction. The phylogenetic roots of these processes might possibly lie much deeper as primates from a wide range of species follow gaze [@rosati2009looking; @tomasello1998five; @itakura2004gaze]. Yet, if they also show the same processing signatures has yet to be explored.

There are important limitations to this study. The methodological factors that influenced performance might have overshadowed individual and community-level differences that originate from other sources. Importantly, this does not affect our interpretation of the data as evidence for shared cognitive processing because the key processing signatures were present in all communities. The role that social interaction plays as a potential driver of development and source of individual differences is vastly under-explored in the current study. The small effect of the exposure to older children does not indicate which aspects of social interaction enhance gaze processing. Naturalistic data directly recording social interactions across communities would offer crucial information to close this explanatory gap. Recent work in the field of language acquisition has shown how technological innovations can be used to close this explanatory gap.

>>>>

```{r fig1, include = T, fig.align = "center", fig.cap = "(A) Data collection sites. Points show approximate geographical location of the data collection sites, coloring shows the sample sizes. (B) Screenshot from the task. The scene depicts the choice phase in a test trial. Participants had to use the gaze of the agent to locate the balloon and click on the hedge where they thought the balloon was. Agents, audio recordings and backgrounds were adpated to each cultural setting. (C) Drawings used as agents across cultural settings. (D) Graphical illustration of the cognitive model. Individuals infer the target of an agent's attention by estimating a vector based on the position of the pupils within the eyes. This process is noisy, illustrated by the different vectors. Individuals differ in their level of precision (indicated by alpha). For a given level of precision, the further out the target lands, the less precise the model predicts individuals to be. Solid and transparent dots show simualted means and individual data points to illustrate the predicted effect of target position.", out.width="75%"}
knitr::include_graphics("../visuals/fig1.png")
```

@@

As the first step, we investigated developmental gains. Across all 17 communities, we found a substantial increase in average levels of precision with age (fixed effect in Bayesian regression model [@burkner2019bayesian]: $\beta$ = `r post_fix%>%filter(term == "ageinyears")%>%pull(Estimate)`, 95% HDI (`r post_fix%>%filter(term == "ageinyears")%>%pull(Q2.5)` - `r post_fix%>%filter(term == "ageinyears")%>%pull(Q97.5)`); range of community-level effects in model: $\beta_{min}$ = `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(lci)` - `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%tail(1)%>%pull(uci)`) to $\beta_{max}$ = `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(lci)` - `r post_agg%>%filter(type == "ageinyears")%>%arrange(mean)%>%head(1)%>%pull(uci)`). Nevertheless, there were also marked differences between communities (see Fig. \@ref(fig:fig2)A). In a cross-validation procedure, in which we trained a regression model on a subset of the data (training data) to later predict the held out data (testing data) [@stengelin2023children], we found that a model that assumed cross-cultural variation to outperform simple models in 100% of cases (see Supplemental Material). At first glance, it seems that highly market-integrated communities around the globe showed higher levels of precision compared to less market-integrated communities (see Tab. \@ref(tab:tab1)). However, believe that these results are better understood in terms of exposure to technology and we discuss them in more detail below. Furthermore, average differences between communities were small compared to differences between individuals: communities did not form homogeneous clusters but largely overlapping distributions in that some individuals from communities with a lower average level of precision performed better compared to some individuals from a setting with a very high average level of precision. Similarly, in all communities, some 4-year-olds outperformed children two years older than them (see Fig. \@ref(fig:fig2)A).

@@

The processing signature was clearly visible across all 17 communities. Precision decreased when the agent looked at locations further away from the center (fixed effect: $\beta$ = `r post_fix%>%filter(term == "targetcentralityx")%>%pull(Estimate)`, 95% HDI (`r post_fix%>%filter(term == "targetcentralitx")%>%pull(Q2.5)` - `r post_fix%>%filter(term == "targetcentralityx")%>%pull(Q97.5)`); range of community-level effects: $\beta_{min}$ = `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(lci)` - `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%tail(1)%>%pull(uci)`) to $\beta_{max}$ = `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(mean)`, 95% HDI (`r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(lci)` - `r post_agg%>%filter(type == "targetcentrality")%>%arrange(mean)%>%head(1)%>%pull(uci)`). Visualization of the data showed the predicted u-shaped pattern in all communities (see Fig. \@ref(fig:fig2)B). These results indicate a universal cognitive process used by children in all communities. There is, however, an alternative way in which the u = shaped pattern might arise: it would also arise when participants ignored the agent's gaze and instead always selected the middle of the screen (center bias). To rule out this alternative explanation, we directly compared three cognitive models that made different assumptions about how participants' responses were generated: the focal vector-based gaze estimation model described above, a center-bias model where participants always select the center, and a random guessing model where participants select random locations. For every community, we found overwhelming support for the gaze estimation model (min $BF_{10}$ > 1000 for comparisons with both alternative models). Taken together, children from all 17 communities processed gaze in similar ways.

@@

On an individual level, other factors likely generated individual differences because there was substantial variation even in communities where almost all children had access to touchscreens. Previous work suggested that social cognition develops in social interaction [@barresi1996intentional; @perner1994theory; @rakoczy2022foundations], in particular, with older siblings [@lewis1996social; @ruffman1998older]. We found tentative support for this idea in the form of a small but consistent effect in the direction that participants living together with more children older than themselves (relative to other children from the same setting) had slightly higher levels of precision ($\beta$ = `r pred_est%>%filter(term == "older_children")%>%pull(Estimate)`, SE = `r pred_est%>%filter(term == "older_children")%>%pull(Est.Error)`, 95% CrI = `r pred_est%>%filter(term == "older_children")%>%pull(Q2.5)` - `r pred_est%>%filter(term == "older_children")%>%pull(Q97.5)`).

Following and understanding gaze is a foundational building block of human social cognition [@tomasello2007reliance; @scaife1975capacity]. A substantial body of work has explored the developmental onset of gaze following in a few selected cultural settings [@moore2008development; @tang2023slow; @gredeback2010development; @byers2021development; @hernik2019infant; @callaghan2011early]. The data reported here provides evidence that children all over the world process gaze in the same way. Key performance signatures of a model that sees gaze following as a form of social vector estimation were found in the data of all 17 distinct cultural communities. The cognitive processes underlying gaze following are thus likely to be rooted in humans' evolved cognitive architecture, which is -- presumably -- later refined during social interaction [@astor2020social; @movellan2002development; @senju2015early]. The phylogenetic roots of these processes might possibly lie much deeper as primates from a wide range of species follow gaze [@rosati2009looking; @tomasello1998five; @itakura2004gaze; @kano2014cross]. Yet, if they also show the same processing signatures has yet to be explored.

There are important limitations to this study. The methodological factors that influenced performance might have overshadowed individual and community-level differences that originate from other sources. Importantly, this does not affect our interpretation of the data as evidence for shared cognitive processing because the key processing signatures were present in all communities. The role that social interaction plays as a potential driver of development and source of individual differences is vastly under-explored in the current study. The small effect of the exposure to older children does not indicate which aspects of social interaction enhance gaze processing. Naturalistic data directly recording social interactions across communities would offer crucial information to close this explanatory gap. Recent work in the field of language acquisition has shown how technological innovations can be used to close this explanatory gap [@bergelson2023everyday; @donnelly2021longitudinal].
